Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-04 docs/manuscript.tex
work on experiment 1

\documentclass[10pt,letterpaper]{article}

@@

Inverse base-rate effect \cite<IBRE, >{medin1988problem} is an irrational tendency in humans to overweigh rare events when faced with ambiguity.
In a traditional design, people learn to categorise two overlapping sets of features under two distinct labels.
These sets share a single feature, A, and posess a unique feature, B and C, predictive of their respective category label.
During learning, these sets of features occur at different frequencies.
The features under the common label usually occur three times as much as features under the rare label \cite{kruschke1996base}.
Following training, people categorise features presented by themselves and un unique combinations.
People optimally label uniquely predictive features, B and C, presented individually with teir respective common and rare label.
Responses on the shared feature A also show base-rate following.
But when uniquely predictive features are paired, B and C, people tend to respond with the rare category label.
According Classical Probability Theory, the rational response is to attribute the common label to this ambiguous compound, because it is the most frequently occuring label.
This rare bias on ambiguous combinations of BC have been observed across different variety of different experiments and manipulations \cite{kalish2001inverse,don2017effects,don2017effects,inkster2022effect,wills2014attention}.
For a more thourough introduction into this irrational bias, see a review by \citeA{don2021attention}.

@@

@@

Recruited 169 participants.

@@

@@

@@

To match performance with

@@

@@

125 ppt made it into the analysis.

\begin{table}[H]
  \begin{center}
    \caption{Stuff. \\}
    \label{tab:results-exp1}
    \vskip 0.12in
    \begin{tabular}{rcc}
      \hline
      & $P(common)$ & $P(rare)$ \\
      \hline
      A & 0.69 & 0.31 \\
      AB & 0.94 & 0.06 \\
      AC & 0.08 & 0.92 \\
      B & 0.94 & 0.06 \\
      \textbf{BC} & \textbf{0.33} & \textbf{0.67} \\
      C & 0.04 & 0.96 \\
    \end{tabular}
  \end{center}
\end{table}

>>>>

\documentclass[10pt,letterpaper]{article}
\usepackage{hyperref}

@@

The \textit{inverse base-rate effect} \cite<IBRE, >{medin1988problem} is an irrational tendency in humans to overweigh rare events when faced with ambiguity.
In a traditional design, people learn to categorise two overlapping sets of features under two distinct labels.
These sets share a single feature, A, and posess a unique feature, B and C, predictive of their respective category label.
During learning, these sets of features occur at different frequencies.
The features under the common label usually occur three times as much as features under the rare label \cite{kruschke1996base}.
Following training, people categorise features presented by themselves and un unique combinations.
People optimally label uniquely predictive features, B and C, presented individually with teir respective common and rare label.
Responses on the shared feature A also show base-rate following.
But when uniquely predictive features are paired, B and C, people tend to respond with the rare category label.
According Classical Probability Theory, the rational response is to attribute the common label to this ambiguous compound, because it is the most frequently occuring label.
This rare bias on ambiguous combinations of BC have been observed across different variety of different experiments and manipulations \cite{kalish2001inverse,don2017effects,don2017effects,inkster2022effect,wills2014attention}.
For a more thourough introduction into this irrational bias, see a review by \citeA{don2021attention}.

@@

Experiment 1 is a conceptual replication of an experiment included in the Appendix of \citeA{johansen2007paradoxical}, but not the main text.
There are no details about the procedure of this experiment, so we cannot make direct comparisons.
The only information available are the list of test items (23), the doubled-up design (2 sets of categories and features), and the sample size.
We substantially simplified our implementation by removing the doubled-up design and reducing the number of test items to 6.

@@

Participants were undergraduate students who recieved course credit for their participantion.
We recruited 169 participants online through the SONA recruitment system.

\subsubsection*{Apparatus}

The experiment was programmed in JsPsych \cite{deleeuw2015JsPsych} to be run in a web browser.
The experiment code is avalaible at \href{https://osf.io/auwvt/?view_only=2dc8384074fa4bcf9f2e3937fdaee2b4}{OSF} and \href{https://github.com/lenarddome/ply216-observational-ibre}{GitHub}.
% links to be inserted here
Participants completed the experiment on their personal computer.
The experiment did not allow the use of tablets and smartphones.

@@

Category labels corresponded with response keys and were called Disease \textbf{Z} and Disease \textbf{L}.
Category features were symptoms: fever, headache, and rash.
These physical features were randomly allocated to abstract features, A, B and C at the beginning of each individual session.
Features and labels appeared in full sentences, such as '\textit{John has fever and rash, which belongs to disease Z}'.
Names were randomly drawn from a pool of male and female first names.
The list were compiled from an online repository of popular baby names\footnote[1]{The list was taken and later curated from \href{https://github.com/aruljohn/popular-baby-names}{a GitHub repository}.}.
We selected the 50 most popular male and female names from 2021.

@@

Table \ref{tab:abstract-exp1} summarises the abstract design of the experiment.
This design is the simplest implementation of the IBRE procedure to date.
Participants completed two phases: a training and test phase.
In the training phase, they encountered descriptions of people, the symptoms they experienced, and their respective disease.
These descriptions appeared in the format of '\textit{John has fever and rash, which belongs to disease Z}'.
Partipants studied these examples and when they were ready to move on, they pressed the spacebar.
They needed to complete reading the description within 5 seconds.
If the 5 seconds threshold has passed, a screen appeared with the message '\textit{Please respond faster!}'.
In each training block, participants encountered 6 common diseases (common category exemplars) and 2 rare diseases (rare category exemplars).
After the second block of training, participants were given a choice.
They could either move straight to the test phase or complete another training block.
There were a maximum of 5 blocks they could complete.

@@

To match performance with the predictive learning implementations of the IBRE, we decided to exclude participants whose test performance on the training items were below 0.75 accuracy.
We arrived at this threshold by testing all difference levels of accuracy through calculating the Bayes factors for binomial data.
We used the method implemented in BayesFactor R package \cite{morey2022bayes}.
If the Bayes Factor fell above 3, we concluded that we have sufficient evidence to believe the participant learned the training items.

@@

In order to test the presence of the IBRE, we calculated a Bayes Factor for a one-sample design.
We calculate the probability of responding rare on the critical BC test item, $P(rare|BC)$, for each participant.
Then we tested this distribution of probabilities against the null, $mu = 0.5$, which denoted random responding.
If the Bayes Factor fell below 1/3, we concluded that participants' responses are not different from random responding.
If the Bayes Factor fell above 3, we concluded that participants' responses reliably differ from null.
If the mean probability of $P(rare|BC)$ is higher then 0.5, we conclude that we observed the IBRE.
Values lower than 0.5 would indicate rational responding.
We used the method implemented in the BayesFactor R package \cite{morey2022bayes}.

@@

After exclusion, 125 participants made it into our main analysis.

\begin{table}[H]
  \begin{center}
    \caption{Group-level mean probabilities for each stimuli presented during the test phase in Experiment 1. \\}
    \label{tab:results-exp1}
    \vskip 0.12in
    \begin{tabular}{rcc}
      \hline
      & $P(common)$ & $P(rare)$ \\
      \hline
      A & 0.69 & 0.31 \\
      AB & 0.94 & 0.06 \\
      AC & 0.08 & 0.92 \\
      B & 0.94 & 0.06 \\
      \textbf{BC} & \textbf{0.33} & \textbf{0.67} \\
      C & 0.04 & 0.96 \\
    \end{tabular}
  \end{center}
\end{table}
